Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-16 _tex/index.tex
Built site for gh-pages

In its most general form, Thurstone's theory
\citeyearpar{Thurstone_1927b} posits that the dichotomous outcome
resulting from comparing two stimuli is determined by two factors: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} refers to the psychological effect each
stimulus has on the judges, or more simply stated, the judges'
perception of the trait level of each stimulus. Thurstone assumes that
the discriminal process for each stimulus follows a Normal distribution.
In this distribution, the mode (mean), known as the \emph{modal
discriminal process}, represents the position of the stimulus on the
trait continuum, while the dispersion, known as the \emph{discriminal
dispersion}, reflects variability in the perceived trait level of the
stimulus. Figure~\ref{fig-discriminal_process} shows example
distributions of discriminal process for two stimuli (objects).

@@

However, since the discriminal mode and dispersion of a single stimulus
are not directly observable without comparing it to other stimuli, the
\emph{law of comparative judgment} becomes essential. This law asserts
that when assessing a specific trait by comparing two stimuli, the
stimulus positioned further along the continuum is perceived as having a
higher level of that trait. Thus, the observed dichotomous outcome is
determined by the distribution of the difference between the discriminal
processes of the two stimuli, called the \emph{discriminal difference}.
Figure~\ref{fig-comparative_judgment} shows an example distribution of
the discriminal difference for two stimuli (objects).

@@

Importantly, the theory's general form primarily addresses pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. Consequently, to enhance its practical
applicability, Thurstone introduced five distinct cases, each defined by
progressively simplifying assumptions. Figure~\ref{fig-thurstone_cases}
summarizes these cases, focusing on key assumptions such as the
distribution of discriminal processes, the similarity of discriminal
dispersion across stimuli, the correlation between stimuli, and which
judges perform the comparisons. For a comprehensive discussion of this
progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248-253]{Bramley_2008}.

@@

\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{./images/tables/thurstone_cases.png}

@@

\caption{\label{fig-thurstone_cases}Thurstones cases and asumptions}

@@

Surprisingly, despite its reliance on the largest number of simplifying
assumptions \citep[pp.~253]{Bramley_2008}, Case V remains the most
widely used case in the CJ literature. This popularity is largely due to
its simplified statistical representation in the Bradley-Terry-Luce
(BTL) model \citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors
Case V's assumptions, with one key difference: while Case V assumes a
Normal distribution for the discriminal processes, the BTL model uses
the more mathematically tractable Logistic distribution
\citep[pp.~254]{Bramley_2008} (see Figure~\ref{fig-thurstone_cases}).
This substitution has little effect on the model's estimation or
interpretation, as the Normal and Logistic distributions differ by a
scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I} (refer to
Figure~\ref{fig-logistic_vs_normal}).

@@

However, Case V (and its BTL model counterpart) was originally developed
to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
over precision in trait measurement. As a result, its assumptions may
not be suitable for applications beyond the psychophysical contexts,
which typically involve simple, homogeneous stimuli. Thurstone himself
cautioned that its use ``should not be made without (an) experimental
test'' \citep[pp.~270]{Thurstone_1927b}, further acknowledging that some
assumptions could become problematic with less homogeneous stimuli, such
as handwriting or English compositions \citep[pp.~374]{Thurstone_1927a}.
Consequently, given that current CJ applications often involve more
complex, less homogeneous stimuli, two assumptions of Case V may not
always hold theoretically or in practice: the assumption of zero
correlation and equal dispersion between stimuli.

{Regarding the assumption of \emph{zero correlation between stimuli}
\ldots{}}

@@

Building on the previous section, it is evident that the BTL model
typically functions as the measurement model for the trait under
assessment \citep{Andrich_1978, Bramley_2008}. A measurement model
specifies how manifest variables contribute to the estimation of latent
variables \citep{Everitt_et_al_2010}. For example, when evaluating text
quality, researchers use the BTL model to process the dichotomous
outcomes resulting from the pairwise comparisons (the manifest
variables) to estimate scores that reflect the underlying quality level
of the texts (the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

>>>>

In its most general form, Thurstone's theory
\citeyearpar{Thurstone_1927b} posits that the dichotomous outcome
resulting from comparing two stimuli is determined by two factors: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} refers to the psychological effect each
stimulus has on the judges, or more simply stated, the judges'
perception of the trait level of each stimulus. Thurstone assumes that
the discriminal process for each stimulus follows a Normal distribution.
In this distribution, the mode (mean), known as the \emph{modal
discriminal process}, represents the position of the stimulus on the
trait continuum, while the dispersion, known as the \emph{discriminal
dispersion}, reflects variability in the stimulus' perceived trait
level. Figure~\ref{fig-discriminal_process} shows example distributions
of discriminal process for two stimuli (objects).

@@

However, since the discriminal mode and dispersion of a single stimulus
are not directly observable except through comparison, the \emph{law of
comparative judgment} becomes essential. This law asserts that when
assessing a specific trait by comparing two stimuli, the stimulus
positioned further along the continuum is perceived as having a higher
level of that trait. Thus, the observed dichotomous outcome is
determined by the distribution of the difference between the stimuli's
discriminal processes, called the \emph{discriminal difference}.
Figure~\ref{fig-comparative_judgment} shows an example distribution of
the discriminal difference for two stimuli (objects).

@@

Importantly, the theory's general form primarily addresses pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. Consequently, to enhance its practical
applicability, Thurstone introduced five distinct cases, each defined by
progressively simplifying assumptions. Table~\ref{tbl-thurstone_cases}
summarizes these cases, focusing on key assumptions such as the
distribution of discriminal processes, the similarity of discriminal
dispersions across stimuli, the correlation between stimuli, and which
judges perform the comparisons. For a comprehensive discussion of this
progression, refer to \citet{Thurstone_1927b} and
\citet[pp.~248-253]{Bramley_2008}.

\begin{table}

\caption{\label{tbl-thurstone_cases}Thurstones cases and asumptions}

@@

\includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{./images/tables/thurstone_cases.png}

@@

\end{table}%

@@

Surprisingly, despite its reliance on the largest number of simplifying
assumptions \citep[pp.~253]{Bramley_2008}, Case V remains the most
widely used case in the CJ literature. This popularity is largely due to
its simplified statistical representation in the Bradley-Terry-Luce
(BTL) model \citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors
Case V's assumptions, with one key difference: while Case V assumes a
Normal distribution for the stimuli's discriminal processes, the BTL
model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Bramley_2008} (see Table~\ref{tbl-thurstone_cases}).
This substitution has little effect on the model's estimation or
interpretation, as the Normal and Logistic distributions differ by a
scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I} (refer to
Figure~\ref{fig-logistic_vs_normal}).

@@

However, Case V (and its BTL model counterpart) was originally developed
to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
over precision in trait measurement. As a result, its assumptions may
not suit applications beyond the psychophysical contexts for which it
was created. Thurstone himself cautioned that its use ``should not be
made without (an) experimental test'' \citep[pp.~270]{Thurstone_1927b},
acknowledging that some assumptions could prove problematic with more
complex, less homogeneous stimuli, such as handwriting or English
compositions \citep[pp.~374]{Thurstone_1927a}. Consequently, given that
current CJ applications often deal with complex, less homogeneous
stimuli, two key assumptions of Case V may not consistently hold in
theory or practice: the zero correlation and equal dispersion between
stimuli.

{The \emph{zero correlation between stimuli} represents two additional
assumtions about the method:}

@@

Building on the previous section, it is evident that the BTL model
typically functions as the measurement model for the trait of interest
\citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
manifest variables contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
researchers use the BTL model to process the dichotomous outcomes
resulting from the pairwise comparisons (the manifest variables) to
estimate scores that reflect the underlying texts' quality level (the
latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
